\begin{acknowledgements}
    I want to first thank my advisor, Jo Bovy, for his guidance and support over the years. I also want to thank my committee members, Suresh Sivanandam and Bob Abraham for their time and feedback.
    
    I am grateful for the support of my friends and family back home, who have remained a stalwart anchor for me through the tumult of graduate school. In particular, I want to thank my parents, Martin and Catherine, who have always believed in me, even when it hasn't been warranted. I would also like to specifically thank my aunt Kate Lane, who has inspired me throughout this process. Among the most important people in my life are my friends from Victoria: Mitch, Adam, Theron, Nick, Cody, Mike. I also want to recognize Carmen and Warren Cabral, who have been welcoming and generous to me over the last three years. Aalbert Van Schothorst deserves special credit for being a valuable teacher and mentor when I was young. I am also glad to have made new friends in Toronto, who I would also like to thank: Steff, Victor, Nathaniel, Phil, Sam, Alex, Shannon, Fraser, Ariel, and many more.
    
    I want to thank professors, mentors, and colleagues who have guided me through my academic career. A special thanks to Helen Kirk, who was my first academic advisor almost 10 years ago. Doug Johnstone, Julio Navarro, Azi Fattahi, Kyle Oman, Steve Mairs, and Mike Chen also mentored me while I was at the University of Victoria. I also have to recognize Kim Venn, who is probably the reason I decided to pursue the stellar halo as a field of study, and deserves much of the thanks for propelling me to graduate school. In Toronto, I have to thank Jason Hunt, Josh Speagle and Ting Li. I also need to give a special thanks to Ted Mackereth, who has been a friend and has acted as my secondary supervisor over many periods throughout my PhD.
    
    I also want to recognize Science communicators who have helped me to keep my head up, and in many respects inspired me to pursue astronomy as a young undergraduate student: Neil deGrasse Tyson, Michael Stevens from Vsauce, Brady Haran from Sixty Symbols, Matt O'Dowd from PBS Space Time, and many others. I began my undergraduate degree in biochemistry, and they are the principal reason I decided to switch my degree to physics, a choice I am glad to have made. Since this is my thesis and I can write whatever I want, I also want to provide my definitive ranking of the top Star Trek characters, who I also want to thank. In order: Picard, Spock, Data, Kirk, McCoy, Worf, Sisko, Seven, Riker, LaForge, Scottie, and Janeway.

    Finally, thanks to Haley. You mean the world to me, and you always knew I could do this.

\end{acknowledgements}